\begin{abstract}
Time series forecasting is fundamental  to data-driven decision-making in domains such as finance, healthcare, manufacturing, and transportation.
%, yet existing methods—including deep learning models (LSTMs, Transformers) and diffusion models—suffer from rigid fixed-length input windows, insufficient multi-scale pattern modeling, and limited spatial-temporal correlation utilization. 
Despite recent progress, existing deep learning approaches--including recurrent architectures, transformer-based models, and diffusion-based frameworks--remain limited by their reliance on fixed-length input windows, inadequate modeling of multi-scale temporal structures, and suboptimal utilization of spatial–temporal dependencies.
To address these limitation, we present the Multi-Resolution Conditional Diffusion Model (MR-CDM), which incorporates three rigorously designed components. 
First, an \textbf{adaptive delay embedding mechanism} maps variable-length time series into structured two-dimensional representations while preserving essential temporal dependencies. 
Second, a \textbf{hierarchical multi-resolution trend decomposition module} captures temporal patterns across distinct scales.
Third, a \textbf{multi-scale conditional diffusion process} integrates contextual information to constrain the generative trajectory and mitigate stochastic variance. 
Comprehensive evaluations on four real-world datasets (Electricity Load, Traffic Flow, Weather and Solar-Energy) demonstrate that MR-CDM  achieves consistent and statistically significant improvements over state-of-the-art baselines (CSDI, TimeGrad, Informer, Autoformer), yielding 5.9–9.9\% reductions in MAE and 6.0–8.9\% reductions in RMSE on average. 
Ablation studies further verify the contribution of each component, with the multi-resolution decomposition producing up to 11.2\% MAE improvement. 
Theoretical analyses establish the statistical consistency of the proposed model, showing that the KL divergence is bounded by $O(1/\sqrt{N})$, and confirm its computational feasibility with a complexity of $O(S \cdot L \log L)$ FLOPs.
These results collectively indicate that MR-CDM provides a principled and effective solution for variable-length, multi-scale time series forecasting.
\end{abstract}